### 2022-03-17 satfol/satfol.tex
more

>>>>

\section{Evaluation}
\label{sec:satfol:evaluation}

% {\bf\large TODO: Evaluate \textit{real} PPE in Figure \ref{fig:preprocess} and discuss SPPE and PPE.
% References to PPE are broken now. }

\newcommand{\baseres}[0]{7897}
% \newcommand{\baseresplusfive}[0]{1563}

\newcommand{\resnum}[1]{
   \FPeval{\result}{clip(#1-\baseres)}%
   \ifnumcomp{\result}{>}{0}{$+\result$}{$\result$}%
}
\newcommand{\resnumphantom}[1]{%
   \FPeval{\result}{clip(#1-\baseres)}%
   \FPeval{\negresult}{clip(-\result)}%
   \ifnumcomp{\result}{>}{0}{\phantom{$0$}{${+}\result$}}{\phantom{$0$}{${-}\negresult$}}%
}
\newcommand{\resnumbold}[1]{
   \FPeval{\result}{clip(#1-\baseres)}%
   \FPeval{\negresult}{clip(-\result)}%
   \ifnumcomp{\result}{>}{0}{\phantom{${+}\result$}\llap{$\mathbf{\boldsymbol{+}\result}$}}{\phantom{${-}\negresult$}$\llap{\mathbf{\boldsymbol{-}\negresult}}$}%
}
\newcommand{\resnumboldphantom}[1]{%
   \FPeval{\result}{clip(#1-\baseres)}%
   \FPeval{\negresult}{clip(-\result)}%
   \ifnumcomp{\result}{>}{0}{\phantom{$0{+}\result$}\llap{$\mathbf{\phantom{0}{\boldsymbol{+}}\result}$}}{\phantom{$0{-}\negresult$}$\llap{\mathbf{\phantom{0}{\boldsymbol{-}}\negresult}}$}%
}
\newcommand{\evaluating}[0]{\textbf{?}}

\looseness=-1
We measure the impact of our elimination techniques for various values of their
parameters. As a baseline, we use Zipperposition's first-order portfolio mode,
which runs the prover in 13~configurations of
heuristic parameters in consecutive time slices. None of these configurations
use our new techniques. To evaluate a given parameter value, we fix
it across all 13~configurations and compare the results with the baseline.

\looseness=-1
The benchmark set consists of all 13\,495 CNF and FOF TPTP 7.3.0 theorems
\cite{gs-17-tptp}.
%
The experiments were carried out on StarExec servers \cite{sst-14-starexec}
equipped with Intel Xeon E5-2609 CPUs clocked at 2.40 GHz. The portfolio mode
uses a single CPU core with a CPU time limit of 180~s. The base
configuration solves \NumberOK{\baseres} problems. The values in the tables
indicate the number of problems solved minus \NumberOK{\baseres}. Thus,
positive numbers indicate gains over the baseline. The best result is shown in
bold.

\ourpara{Hidden-Literal-Based Elimination}
% 
The first experiments use all implemented HLBE rules.
To avoid overburdening Zipperposition, we can enable an option to limit the
number of tracked clauses for hidden literals. Once the limit has been reached, any
request for tracking a clause will be rejected until a tracked
clause is deleted. We can choose which kind of clauses are
tracked:\ only clauses from the active set $\mathcal{A}$,
only clauses from the passive set $\mathcal{P}$, or both.
We also vary the maximal implication chain length $K_\mathrm{len}$ and the
number of computed self-implications $K_\mathrm{imp}$.

In Zipperposition, every lookup for instances or generalizations of
$\eqlit{s}{t}$ must be done once for each orientation of the equation.
To avoid this inefficiency, and also because the implementation of hidden literals does not fully
exploit congruence, we can disable\pagebreak[2] tracking clauses with at least one
functional literal. Clauses containing functional literals can then still be
simplified.

Figures~\ref{fig:hle-clauses-neq} and \ref{fig:hle-clauses-eq}
%~\ref{fig:hle}
show the results, without and with functional literal tracking enabled,
for $K_\mathrm{len} = 2$ and $K_\mathrm{imp} = 0$.
The columns specify different limits on the number of tracked clauses, with $\infty$
denoting that no limit is imposed. The rows represent different kinds of tracked clauses.
The results suggest that tracking functional literals is \NumberOK{not} worth the effort
but that tracking predicate literals \NumberOK{is}. The best improvement is observed when
both active and passive clauses are tracked. Normally DISCOUNT-loop
provers \cite{adf-1995-discount} such as Zipperposition do not simplify
active clauses using passive clauses, but here we see that this can be effective.
%
%--Raw data shows that when up to 250 clauses in either set are
%--tracked, tracking functional literals helps solve \NumberNOK{5} problems that
%--cannot be solved by a configuration that tracks only predicate literals or the
%--base configuration. This suggests that a more optimized indexing implementation
%--would help our technique reach its full potential.
Figure~\ref{fig:km-kimp} shows the impact of varying $K_\mathrm{len}$ and
$K_\mathrm{imp}$, when 500 clauses from the entire proof state are tracked.
These results suggest that computing long implication chains \NumberOK{is counterproductive.}
% Furthermore, raw data reveals that there is
% no gain in this case: not a single problem that configuration with $K_\mathrm{len}=2$ and
% $K_\mathrm{imp}=0$ cannot solve is solved when $K_\mathrm{imp}K_\mathrm{len}$ is set to 8 and
% $K_\mathrm{imp}$ to 0.

\begin{figure}[t]
   \centering
   \begin{minipage}{.48\textwidth}
     \begin{center}
       \def\arraystretch{1.1}%
       \relax{\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
         & \multicolumn{4}{c}{Tracked clauses} \\
         & $250$        & $500$         & $1000$       &$\infty$ \\ \midrule 
\strut Active       & \resnum{7883}  & \resnum{7881}   & \resnum{7889}  & \resnum{7885}    \\
\strut Passive      & \resnumphantom{7904}  & \resnum{7907}   & \resnum{7902}  & \resnum{7862}    \\
\strut Both         & \resnumbold{7909}  & \resnum{7907}   & \resnum{7904}  & \resnum{7852} \\     
\end{tabular}}
\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when only predicate literals are tracked}
\label{fig:hle-clauses-neq}
     \end{center}
   \end{minipage}
   \hfill
   \begin{minipage}{.45\textwidth}
     \begin{center}
       \def\arraystretch{1.1}%
       \relax{\begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1em}c@{\hskip1em}c@{\hskip 1em}c}
         & \multicolumn{4}{c}{Tracked clauses} \\
         & $250$                   & $500$                 & $1000$                    & $\infty$ \\ \midrule 
\strut Active       & \resnum{7887}            & \resnum{7883}         & \resnumphantom{7889}      & \resnum{7879}    \\
\strut Passive      & \resnumphantom{7892}     & \resnumphantom{7892}  & \resnum{7883}             & \resnum{7826}    \\
\strut Both         & \resnumboldphantom{7899} & \resnumphantom{7896}  & \resnumphantom{7889}      & \resnum{7818} \\ 
\end{tabular}}
\caption{Impact of the number and kinds of tracked clauses on HLBE performance, when all literals are tracked}
\label{fig:hle-clauses-eq}
     \end{center}
   \end{minipage}
 \end{figure}

\begin{figure}[t]
\small
\centering
\def\arraystretch{1.1}%
\begin{tabular}[t]{l@{\hskip 2em}c@{\hskip 1.5em}c@{\hskip1.5em}c@{\hskip 1.5em}c} 
                             & \multicolumn{4}{c}{Chain length $K_\mathrm{len}$} \\
                             & $1$                   & $2$                 & $4$                   & $8$ \\ \midrule 
   $K_\mathrm{imp}=0$        & \resnum{7906}         & \resnum{7907}       & \resnum{7904}         & \resnum{7902}    \\
   $K_\mathrm{imp}=1$        & \resnum{7902}         & \resnumbold{7908}       & \resnum{7904}         & \resnum{7901}    \\
   $K_\mathrm{imp}=2$        & \resnum{7903}         & \resnumbold{7908}       & \resnum{7905}         & \resnum{7905}    \\
\end{tabular}
\caption{Impact of the parameters $K_\mathrm{len}$ and $K_\mathrm{imp}$ on HLBE performance}
\label{fig:km-kimp}
\end{figure}

\begin{figure}[t!]
   \small
   \centering
   \def\arraystretch{1.1}%
   \begin{tabular}[t]{l@{\hskip 1.5em}c@{\hskip 1.5em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c@{\hskip 1em}c}
                           &               & \multicolumn{5}{c}{Relaxed with $K_\mathrm{tol}$} \\
                           & K\&K          & $0$             & $25$           & $50$           & $100$    & $200$    \\ \midrule 
   \infname{SPE} preproc.  & \resnum{7967} & \resnum{8014}   & \resnum{8051}  & \resnum{8057}  & \resnum{8051}      & \resnum{8055} \\
   \infname{PPE} preproc.  & \resnum{7968} & \resnum{8021}   & \resnum{8057}  & \resnum{8061}  & \resnumbold{8062}      & \resnum{8059} \\
   \end{tabular}
   \caption{Impact of the choice of criterion on predicate elimination performance}
   \label{fig:criterion}
\end{figure}

\begin{figure}[t]
   \small
   \centering
   \def\arraystretch{1.1}%
   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c}
                         &               &               &                &               &                    & \smash{HLBE} \\[-1pt] %% TYPESETTING
                         &               &               & \smash{SPE}    &               & \smash{PPE}        & \smash{+PPE} \\[-1pt] %% TYPESETTING
                         & BCE           & SPE           & +BCE           & PPE           & +BCE               & +BCE           \\ \midrule 
   \strut Preprocessing  & \resnum{7927} & \resnum{8051} & \resnum{8056}  & \resnum{8057} & \resnumbold{8063}  & \resnum{8059} \\
   \strut Inprocessing   & \resnum{7849} & \resnum{8037} & \resnum{8024}  & \resnum{8043} & \resnum{8028}      & \resnum{8024}     \\
   \end{tabular}
   \caption{Performance of predicate and blocked clause elimination}
   \label{fig:preprocess}
\end{figure}

\renewcommand{\baseres}{856}

\begin{figure}[t]
   \small
   \centering
   \def\arraystretch{1.1}%
   \begin{tabular}[t]{l@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1.3em}c@{\hskip 1em}c@{\hskip 1em}c} \\
                         &              &              &                   &              &              & \smash{HLBE}   \\[-1pt] 
                         &              &              & \smash{SPE}       &              & \smash{PPE}  & \smash{+PPE}   \\[-1pt] 
                         & BCE          & SPE          & +BCE              & PPE          & +BCE          & \smash{+BCE}    \\ \midrule 
   \strut Preprocessing  & \resnum{885} & \resnum{902} & \resnumbold{916}  & \resnum{903} & \resnum{915} & \resnum{911}  \\
   \end{tabular}
   \caption{Performance of predicate and blocked clause elimination for establishing satisfiability}
   \label{fig:bce-sat}
\end{figure}

\ourpara{Predicate and Blocked Clause Elimination}
%
% \textit{Evaluation results discussion paragraph---here effects of definitions can be discussed}
% \textit{Do not forget to mention SPE was already implemented and }
For defined predicate elimination, the number of resolvents grows
exponentially with the number of occurrences of $\cst{p}$. To avoid this expensive
computation, we limit the applicability of PPE to proof states for which $\cst{p}$
is singular. According to our informal experiments, full PPE,
without this restriction, generally performs less well.

\begin{qle}\color{red} (TODO: Add QLE)\color{black}\end{qle}

Predicate elimination can be done using Khasidashvili and Korovin's
criterion (K\&K) or using our relaxed criterion with different values of
$K_\mathrm{tol}$. Figure~\ref{fig:criterion} shows the results
for SPE and PPE used as preprocessors.
Our numbers corroborate Khasidashvili and Korovin's findings:
SPE with K\&K proves \NumberOK{70} more problems than the base, a
\NumberOK{0.9\%} increase, comparable to the 1.8\% they observe when
they combine SPE with additional preprocessing.
Remarkably, the number of additional proved problems more than
\NumberOK{doubles} when we use our criterion with $K_\mathrm{tol} > 0$, for both SPE and PPE.

\looseness=-1
Although this is not evident in Figure~\ref{fig:criterion}, varying
$K_\mathrm{tol}$ substantially changes the set of problems solved. For example,
when $K_\mathrm{tol}=0$, SPE proves \NumberOK{60} theorems not proved using
$K_\mathrm{tol}=50$. The effect weakens as $K_\mathrm{tol}$ grows. When
$K_\mathrm{tol}=100$, SPE proves only \NumberOK{13} problems not found when
$K_\mathrm{tol}=200$. Similarly, the set of problems proved by SPE and PPE
differs: When $K_\mathrm{tol}=25$, \NumberOK{14} problems are proved by PPE but
missed by SPE. Recognizing definition sets is useful:
PPE \NumberOK{outperforms} SPE regardless of the criterion.

% \looseness=-1 Having implemented those techniques, we are in position to answer
% two interesting questions: {How helpful are definition clauses in defined
% predicate elimination (DPE)?} and {What are the effects of performing blocked
% clause elimination (BCE) and DPE until fixpoint (BD)?}. The second question was
% also of interest to Kiesl et al.\ \cite{ksstb-2017-blockedfol}, but was left
% unanswered as they did not implement predicate elimination.

\looseness=-1
Performing BCE and variable elimination until fixpoint increases the performance
of SAT solvers~\cite{jbh-10-BCE}. We can check whether the same holds for superposition provers.
In this experiment, we use the relaxed criterion with
$K_\mathrm{tol}=25$ and HLBE which tracks up
to 500 clauses from any clause set, $K_\mathrm{len}=2$, and
$K_{\mathrm{imp}}=0$. We use each technique as preprocessing and inprocessing.

The results are summarized in Figure~\ref{fig:preprocess}, where
the +~sign denotes the combination of techniques.
We confirm the results obtained by Kiesl et al.\ about the performance of BCE
as preprocessing:
It helps prove \NumberOK{30} more problems from our benchmark set, increasing the
success rate by roughly \NumberOK{0.4\%}. The same percentage increase was obtained Kiesl et al.
%
Using BCE as inprocessing, however, hurts performance, presumably because of
its incompatibility with the redundancy criterion.

\looseness=-1%
For preprocessing, the combinations SPE+BCE and PPE+BCE performed roughly \NumberOK{on a
par} with SPE and PPE, respectively. This stands in contrast to the situation
with SAT solvers, where such a combination usually helps. It is also worth noting
that the inprocessing techniques never outperform their preprocessing
counterparts.
%
The last column shows that combining HLBE with other elimination techniques
\NumberOK{overburdens} the prover.

\ourpara{Satisfiability by Blocked Clause Elimination}
%
\looseness=-1%
Kiesl et al.\ found that blocked clause elimination is especially effective on
satisfiable problems. To corroborate their results and ascertain
whether a combination of predicate elimination and blocked clause elimination
increases the success rate, we evaluate BCE on all 2273
satisfiable or TPTP FOF and CNF problems.
The hardware and CPU time limits are the same as in the experiments above.
Figure~\ref{fig:bce-sat} presents the results.

\looseness=-1
The baseline establishes the satisfiability of \baseres~problems. We consider
only preprocessing techniques, since BCE compromises refutational
completeness---a saturation does not guarantee that the original problem was
satisfiable. We note that recognizing definition sets makes almost no difference on
satisfiable problems. The sets of problems solved by BCE and PPE
differ---\NumberOK{30} problems are solved by BCE and not by PPE.
% Kiesl sees around 2.5%, we see 3.5% with BCE

\section{Discussion and Related Work}
\label{sec:satfol:discussion-and-related-work}

We briefly surveyed related work in Sect.~\ref{sec:satfol:introduction}. In this
section, we give a more detailed overview and further discuss connections with
the related work.

The research presented in this \paper{} is two-pronged. For SAT elimination
techniques already generalized to preprocess first-order problems, we looked for
ways to interleave them with the given clause procedure of a superposition
prover, as inprocessing. For techniques that had not yet been ported to
first-order logic, we looked for generalizations that allow both preprocessing and
inprocessing.

Hidden tautology elimination was first described
by Heule et al.\ \cite{hjb-2010-cl-elim}. Better implementation that also
supports hidden literal elimination was later described by the same group of
authors \cite{hjb-2011-big-simplification}. We generalized the underlying
theoretical concepts to first-order logic, and provided an efficient way to deal
with the infinite number of hidden literals that arise  with this
generalization. More efficient graph-based techniques are yet to be explored.

Variable elimination, based on Davis--Putnam resolution \cite{dp-60-dp}, has been studied in
the context of both propositional logic \cite{sp-04-niver,cs-00-zres} and QBF
\cite{ab-2004-re}. It was generalized to first-order logic (as a
preprocessor) by Khasidashvili and Korovin \cite{kk-2016-pe-fol} yielding a technique called
predicate elimination. Improvement of variable elimination,
which uses formula definition information, has been popularized as a preprocessing and
inprocessing technique for CDCL solvers by E{\'{e}}n and Biere
\cite{eb-2005-satpreprocess}. We generalized this improvement to first-order
logic and combined it with existing extension. With tolerable restrictions,
this extension can be used as an inprocessing technique. In SAT and QBF, it was observed that
allowing variable elimination to slightly increase in the clause set size
improves performance \cite{bls-11-bloqqer}. We implemented a similar approach,
achieving the double the number of additional proofs found compared to more
restrictive approaches.

Blocked clause elimination is used in both SAT~\cite{jbh-10-BCE} and QBF
solvers \cite{bls-11-bloqqer}. Its generalization to first-order logic
\cite{ksstb-2017-blockedfol} has showed positive effects when used as a
preprocessor. We showed that blocked clauses cannot be removed during
saturation, but that they can be effectively used to show satisfiability of the
clause set. A combination of blocked clause elimination and
variable elimination performs well in propositional logic \cite{jbh-10-BCE}, but we observed no substantial
improvement when their generalizations are combined.

Our approach is one of many ways to combine ideas from SAT solving and
first-order proving. Other noteworthy architectures that either incorporate a
SAT solver or that generalize the CDCL calculus include DPLL($T$) with
quantifier instantiation
\cite{de-moura-bjoerner-2007,barbosa-et-al-2017,reynolds-et-al-2018},
DPLL($\mupGamma+T$) \cite{bonacina-et-al-2009},
labeled splitting \cite{fietzke-weidenbach-2009},
AVATAR \cite{av-2014-avatar},
MCSAT \cite{de-moura-jovanovic-2013},
CDSAT \cite{bonacina-et-al-2017}, and
SGGS~\cite{bonacina-plaisted-2014}.

\section{Conclusion}
\label{sec:satfol:conclusion}

We adapted several preprocessing and inprocessing elimination techniques implemented in
modern SAT solvers so that they work in a superposition prover. This involved lifting
the techniques to first-order logic with equality but also tailoring them to
work in tandem with superposition and its redundancy criterion.
Although SAT solvers and superposition provers embody radically different
philosophies, we found that the lifted SAT techniques provide valuable optimizations.

\pagebreak[2]
We see several avenues for future work. First, the implementation of hidden
literals could be extended to exploit equality congruence. Second, although
inprocessing blocked clause elimination is incomplete in general, we hope to achieve
refutational completeness for a substantial fragment of it. Third, predicate and
blocked clause elimination, which thrives on the absence of clauses from the
proof state, could be enhanced by tagging and ignoring generated clauses that
have not yet been used to subsume or simplify untagged clauses. Fourth,
predicate and blocked clause elimination could be extended to work with
functional literals. Fifth, more SAT techniques could be adapted, including
bounded variable addition~\cite{mhb-12-reencoding} and blocked clause
addition~\cite{ok-99-er}. Sixth, the techniques we covered could be adapted to
work with other first-order calculi, or generalized
further to work with higher-order calculi such as combinatory superposition
\cite{br-20-full-sup-w-combs} and \osup{}.
